Playing with PCAs

### 2023-12-06 3-reports/playing_with_pcas.Rmd
fix PCA nomenclature. plot PC1

data_combined_subset = make_data_subset(data_combined) %>% filter(!analysis %in% "LOI")
data_combined_wide = make_data_wide(data_combined_subset, sample_key)
data_combined_wide_NO_DIC = make_data_wide(data_combined_subset %>% 
                                             filter(!grepl("dic", name, ignore.case = TRUE)),
                                           sample_key)
```

@@

data_wide_PCA = data_combined_wide %>% dplyr::select(-c("dic", "gwc"))

@@

  ggbiplot(pca_overall$pca_int, obs.scale = 1, var.scale = 1,
           groups = as.character(pca_overall$grp$transect), 
           ellipse = TRUE, circle = FALSE, var.axes = TRUE, alpha = 0) +
  geom_point(size=3,stroke=1, alpha = 1,
             aes(shape = pca_overall$grp$region,
               color = pca_overall$grp$transect))+ 
 # scale_color_manual(values = c("#16879C", "#BB281E" ))+
  #scale_shape_manual(values = c(21, 19))+
  #scale_shape_manual(values = c(21, 21, 19), name = "", guide = "none")+
  # xlim(-4,20)+
  # ylim(-8,8)+
  labs(shape="",
       title = "Overall PCA, both regions",
       subtitle = "Surface horizons only")+
  theme_kp()+
  NULL

@@

pca_overall_upland = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "upland"))
pca_overall_transition = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "transition"))
pca_overall_wetland = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "wc"))

@@

pca_upland_wle = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "upland" & region == "WLE"))
pca_upland_cb = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "upland" & region == "CB"))

@@

pca_transition_wle = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "transition" & region == "WLE"))
pca_transition_cb = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "transition" & region == "CB"))

@@

pca_wetland_wle = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "wc" & region == "WLE"))
pca_wetland_cb = fit_pca_function(data_combined_wide_NO_DIC %>% filter(transect == "wc" & region == "CB"))

@@

>>>>

data_combined_subset = make_data_subset(data_combined) %>% filter(!analysis %in% "LOI")
data_combined_wide_NO_DIC = make_data_wide(data_combined_subset %>% 
                                             filter(!grepl("dic", name, ignore.case = TRUE)),
                                           sample_key)
```

@@

data_combined_wide = make_data_wide(data_combined, sample_key)
data_wide_PCA = 
  data_combined_wide %>% 
  dplyr::select(-c("dic", "gwc", "percentOM")) %>% 
  rename(CEC = cec,
         NH4N = nh4n,
         NO3N = no3n,
         P = mehlichp,
         WEOC = npoc)

@@

#  ggbiplot(pca_overall$pca_int, obs.scale = 1, var.scale = 1,
#           groups = as.character(pca_overall$grp$transect), 
#           ellipse = TRUE, circle = FALSE, var.axes = TRUE, alpha = 0) +
#  geom_point(size=3,stroke=1, alpha = 1,
#             aes(shape = pca_overall$grp$region,
#               color = pca_overall$grp$transect))+ 
# # scale_color_manual(values = c("#16879C", "#BB281E" ))+
#  #scale_shape_manual(values = c(21, 19))+
#  #scale_shape_manual(values = c(21, 21, 19), name = "", guide = "none")+
#  # xlim(-4,20)+
#  # ylim(-8,8)+
#  labs(shape="",
#       title = "Overall PCA, both regions",
#       subtitle = "Surface horizons only")+
#  theme_kp()+
#  NULL

@@

pca_overall_upland = fit_pca_function(data_wide_PCA %>% filter(transect == "upland"))
pca_overall_transition = fit_pca_function(data_wide_PCA %>% filter(transect == "transition"))
pca_overall_wetland = fit_pca_function(data_wide_PCA %>% filter(transect == "wc"))

@@

pca_upland_wle = fit_pca_function(data_wide_PCA %>% filter(transect == "upland" & region == "WLE"))
pca_upland_cb = fit_pca_function(data_wide_PCA %>% filter(transect == "upland" & region == "CB"))

@@

pca_transition_wle = fit_pca_function(data_wide_PCA %>% filter(transect == "transition" & region == "WLE"))
pca_transition_cb = fit_pca_function(data_wide_PCA %>% filter(transect == "transition" & region == "CB"))

@@

pca_wetland_wle = fit_pca_function(data_wide_PCA %>% filter(transect == "wc" & region == "WLE"))
pca_wetland_cb = fit_pca_function(data_wide_PCA %>% filter(transect == "wc" & region == "CB"))

@@

## Plotting PC1 (drivers of variability)

```{r}
get_pc_values = function(dat){

  pca = fit_pca_function(dat)

  x = pca$pca_int$rotation
  x2 = 
    x %>% 
    as.data.frame() %>% 
    dplyr::select(PC1) %>% 
    arrange(desc(PC1)) %>% 
    rownames_to_column("variable") %>% 
    mutate(variable = fct_reorder(variable, (PC1)))

  x2
}
```

```{r, fig.height=8, fig.width=6}

pc_overall = get_pc_values(data_wide_PCA) %>% mutate(region = "combined")
pc_wle = get_pc_values(data_wide_PCA %>% filter(region == "WLE")) %>% mutate(region = "WLE")
pc_cb = get_pc_values(data_wide_PCA %>% filter(region == "CB")) %>% mutate(region = "CB")

pc_all = 
  bind_rows(pc_overall, pc_wle, pc_cb) %>% 
  mutate(variable = fct_reorder(variable, (PC1)))

pc_all %>% 
  mutate(variable = fct_reorder(variable, (PC1))) %>% 
  filter(region != "combined") %>% 
  ggplot(aes(x = PC1, y = variable, color = region))+
  geom_point(size = 5)+
  scale_colour_manual(values = c("#81C0A7", "#1A421D"))+
  xlim(-0.4, 0.4)+
  geom_vline(xintercept = 0)+
  labs(y = "")+
  theme_kp()

pc_all %>% 
  mutate(variable = fct_reorder(variable, (PC1))) %>% 
  filter(region != "combined") %>% 
  ggplot(aes(x = abs(PC1), y = variable, color = region))+
  geom_point(size = 5)+
  scale_colour_manual(values = c("#81C0A7", "#1A421D"))+
 # xlim(0, 0.4)+
  #geom_vline(xintercept = 0)+
  labs(y = "",
       x = "|PC1|")+
  theme_kp()
